Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
adding context and content

% Below: What are the goals of this discussion? TODO: Better motivate. May be make a bried list of items discussed in the workshop.
The discussion is aimed at understanding the practicalities of the key points
discussed in the workshop, and we hope to extract the best practices and
associated with the case studies. We discuss how the cases are applicable to
and have impacted user communities over their lifetime, isolating similarities
and differences between projects. We discuss the takeaways for new
projects and the lessons learned. We try to understand how the course of current
and future development projects might be altered to implement lessons learned.

\subsection{Utility Software}
Utility software has the potential of a broader user base. This is especially
relevant in the modern era of social computing. In the context of research, a
utility software project could be defined as general purpose system applicable
to one or many generic tasks and/or enabling other software to run by providing
a suitable environment. Examples are collaborative development frameworks such
as \emph{GitHub} and \emph{Bitbucket}, distributed and workflow computing
frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and
\emph{VisTrails}, and visualization frameworks such as \emph{VisIT}. Often
projects start with a specific, small set of requirements and users in mind but
expand to provide functionality and appeal to a wider user community. Owing to
its generic nature, utility software finds a wider user community which in turn
leads to a higher participation in development.  Consequently, the process
becomes user-driven and self-sustaining.

Software developed in such a scenario often benefits from
community driven development, deployment and maintenance best practices. One
such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standard practices such as test-driven
development and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}.

Utility software is often developed aimed at better utilizing particular, new
infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
Development in many cases is often a high risk/reward undertaking and explores
uncharted territories.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?  If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.  Something in the framing of this subsection doesn't feel right to me.  FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure. Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code. Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind. New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see many submissions around them in the workshop.}
%
%
Successful projects reap high rewards and have longer usage span. Long
requirement-development-test cycles.  Wide community support but tends to be
handled by specialists rather then end users. Promotes collaborations across
the breadth (e.g., different departments) and depth (e.g., stakeholders within
a department) of community, one of the key ingredient of a sustainable process.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
associated with software projects dealing with new architectures. To leverage
the power of accelerators such as GPUs and MICs new code and libraries are
required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
met with a limited success but nonetheless with many invaluable lessons were
learned about influential cultural and technical aspects in sustainable
software development practices.

Another paradigm in utility software is the software delivered as service over
the web. With increasing popularity of cloud-based computational environments,
many users are leaning towards tools used as service. \emph{GitHub} and
\emph{Bitbucket} can be arguably considered to be such tools, catering to
collaborative development. For scientific users \emph{Globus}-based tools are
a case of service-based utility software discussed during the workshop. The
data movement, authentication and sharing services offered by Globus can be
easily used over the web by collaborating researchers.

\subsection{Scientific software}
Scientific software consists primarily of special-purpose software that was
developed for a target use-case scenario or fixed/frozen requirements in mind.
Software projects pertaining to specific scientific domain often tend to be in
a niche and the user community tends to be small to medium.
%
Specific needs such as numerical accuracy and algorithmic optimization are some
of the paramount requirements of most scientific software.
%
The resulting codebase is relatively small. A smaller codebase and fixed
requirements result in stability, ease of installation, and configurability.
Many such projects are treated as libraries to be used into larger systems such as
some of the utility software discussed in the previous section.
%
While the
software can stay stable and require relatively low maintenance, the
responsibility is often on the shoulders of a very few developers.
%
Community
participation in the development tends to be low. Development is linear and
simplistic with a limited scope to follow software best practices.
%
Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
Most of the above efforts are focused on one scientific and/or applied
mathematics domain. One exception was \emph{Kitware}, which while being a
software product specializing in the scientific process, has a core focus of
developing communities around software processes. An instance of this process
is the development of the \emph{CMake} build utility, which started out as a
building tool for \emph{ITK} but grew to become a generic build utility for C++
projects.
%
Sustainability of scientific research software is often a significant
challenge. Many submissions reported that software is practically considered a
`byproduct' of the actual research. Others contended that the software was not
the main funded part of their research.

%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?

>>>>

% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
We aim to understand the practicalities of the key points discussed in the
workshop, and hope to extract the best practices associated with the case
studies. We discuss how the issues are applicable to the cases and try to
understand how the course of current and future development projects might be
altered to implement lessons learned. In particular, we discuss the following issues:
\begin{itemize}
    \item How closely are the best practices discussed in the workshop were followed?
    \item Career path: Are the original developers still around?
    \item What conscious sustainability efforts were undertaken by stakeholders?
    \item What is the dominant sustainability factor: science, publication, (re)usability?
\end{itemize}

\subsection{Utility Software}

We classify software developed with a potentially wider audience and general
purpose usage in mind as Utility software. In the context of research, a
utility software could be defined as general purpose system applicable to one
or many generic tasks and/or enabling other software to run by providing a
suitable environment. Utility software typically do not address a fundamental
research challenge in a given scientific domain. Examples are collaborative
development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
and workflow computing frameworks such as \emph{Galaxy}, \emph{Swift},
\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
\emph{VisIT}.

% What are the sustainability challenges? What are the oppurtunities?
Development is often a high risk/reward undertaking and explores uncharted
territories and is largely dominated by (re)usability factors. Owing to
relatively large number of features, the development and prototype process is
also lengthy which poses a significant survival risk.
Ferenbaugh~\cite{Ferenbaugh_WSSSPE} further discusses the risks and challenges
associated with software projects dealing with new architectures.

On the other hand, owing to its generic nature, utility software presents
oppurtunities to be usable by a larger community. This makes it more visible
which in turn leads to a broader and deeper participation. For
instance, promoting collaborations across the breadth (e.g. different
departments) and depth (e.g. stakeholders within a department) of community,
one of the key ingredient of a sustainable process. Successful projects reap
high rewards and have longer usage span. Development process becomes
user-driven and self-sustaining.

One such example is the \emph{Galaxy} project. It follows agile software
development practices and implements standard practices such as test-driven
development and socialized bug managing practices via \emph{trello}. Galaxy
\emph{histories} and \emph{toolshed} offers easy community sharing of data and
tools further promoting a collaborative environment. The project very closely
follows the guidelines described in~\cite{Carver_WSSSPE} and many
from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
aiming better utilizing a particular, new infrastructure and architecture,
e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
emerging architectures such as CPU accelerators new code and libraries are
required. The experience of one such effort as described
in~\cite{Ferenbaugh_WSSSPE} met with a limited success but nonetheless with
many invaluable lessons were learned about influential cultural and technical
aspects in sustainable software development practices.

A relatively new paradigm in utility software is the software delivered as
service over the web. With increasing popularity of cloud-based storage and
computational environments, many users are leaning towards tools used as
services. \emph{GitHub} and \emph{Bitbucket} can be arguably considered to be
such tools, catering to collaborative development. For scientific users
\emph{Globus}-based tools are a case of service-based utility software
discussed during the workshop. The data movement and sharing services offered
by Globus can be easily used over the web by collaborating researchers.
%
%
%\note{Dan: so is the only difference here that this category is aimed at new
%systems, where utility software is not?  If so, perhaps this should be
%incorporated into that subsection, so there would just be two subsections.
%Something in the framing of this subsection doesn't feel right to me.  FL: I
%respectfully disagree the the only reason to develop frameworks is to better
%utilize (new or old) infrastructure. Another really important reason is to
%modularize development, making it much easier for an individual developer/user
%to integrate their own code. Ketan: I think in this section I wanted to bring
%up a class of software projects which are undertaken with the main aim of
%tackling new systems and architectures in mind. New infrastructures such as
%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
%many submissions around them in the workshop.}
%
%

\subsection{Scientific software}

Scientific software consists primarily of special-purpose software that was
purpose-built for a target use-case scenario or fixed/frozen requirements in
mind.  Software projects pertaining to specific scientific domain often tend to
be in a niche and the user community tends to be small to medium. They are
mostly driven by the science and specific needs of a research group.

% What are the challenges?
Long-term sustainability of scientific research software is often a significant
challenge. Many submissions reported that software is practically considered a
`byproduct' of the actual research. Others contended that the software was not
the main funded part of their research.  Specific needs such as numerical
accuracy and algorithmic optimization are some of the paramount requirements of
most scientific software.  The resulting codebase is relatively small. A
smaller codebase and fixed requirements result in stability, ease of
installation, and configurability.  Many such projects mature and are treated
as libraries to be used into larger systems such as some of the utility
software discussed in the previous section.  While the software can stay stable
and require relatively low maintenance, the responsibility is often on the
shoulders of a very few developers who might not be specialists in software
development. Development tends to be linear and simplistic with a limited
scope to follow software best practices.

% TODO: What are the oppurtunities?

Some examples of such software discussed as part of the workshop are,
\emph{DUNE}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
Most of the above efforts are focused on one scientific and/or applied
mathematics domain. However, sometimes such projects grow beyond the initial
vision of developers. One such example is \emph{Kitware}, which while being a
software product specializing in the scientific process, has a core focus of
developing communities around software processes. An instance of this process
is the development of the \emph{CMake} build utility, which started out as a
building tool for \emph{ITK} but grew to become a generic build utility for C++
projects.
%

%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
It appears that the career paths for utility software developers are broader
compared to the developers of niche software. However, as projects as a whole,
the small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. The development of scientific
software requires more scientific skills compared with those of utility
software. This diminishes the need of specialist programmers and practices in
many cases where the bulk of development is carried out by a domain scientist.
